---
title: "Ploidy Analysis"
author: "Your Name"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning=FALSE
  
)

library("tidyverse")
library("patchwork")
```

```{r load_data, include=FALSE}
# Load project settings
source(here::here("settings.R"))

# load output functions
source(file.path(functions.dir, "output.R"))

# load design
source(file.path(data.raw.dir, "design.R"))
df.design <- load_design()

# load cell extracts
df.cell.extracts <- read_output_csv("cell_extracts")

# load qPCR data
df.qpcr <- read_output_csv("qpcr", "qpcr_fragments_(20021[2-3]_1)")
```

Calculate the ploidy level using the qpcr and cell extract information.

```{r}
df.ploidy <- df.qpcr %>% 
  left_join(df.cell.extracts, by = "sample_id") %>% 
  mutate(
    ploidy = (fragments_sample_uL * 1E3) / broken_cells_mL
  ) %>% 
  left_join(df.design, by = "sample_id")
```

Inspect the raw data, by plotting the ploidy level per sample and per plate.

```{r}
df.ploidy %>%
  ggplot(aes(x = sample_id, y = ploidy, fill = plate_id, group = plate_id)) +
  stat_summary(fun.y = mean, geom = "bar", position = "dodge") +
  stat_summary(fun.data = mean_se, geom = "errorbar", position = position_dodge(width=0.9)) +
  coord_flip() +
  expand_limits(y=0) +
  labs(
    x = "Sample",
    y = "# Chromosomes per Cell",
    fill = "Plate ID"
  )
```

```{r}
p.before <- df.ploidy %>%
  ggplot(aes( x = sample_id, y = ploidy, colour = as.factor(dilution_factor), shape = purpose)) +
  facet_wrap(~purpose, scales = "free_y", ncol = 1) +
  geom_jitter(width = 0.1) +
  coord_flip()

p.before
```

Are we content with the data? Should we remove some data?

```{r}
df.ploidy.f <- df.ploidy %>%
  filter(plate_id == "200213_1") %>%
  filter(!(sample_id == "200206_2_8" ))
```

Now repeat the last plot to inspect the changes.

```{r}
df.ploidy.f %>%
  ggplot(aes(x = sample_id, y = ploidy, fill = purpose)) +
  stat_summary(fun.y = mean, geom = "bar", position = "dodge") +
  stat_summary(fun.data = mean_se, geom = "errorbar", position = position_dodge(width=0.9)) +
  coord_flip() +
  expand_limits(y=0) +
  labs(
    x = "Sample",
    y = "# Chromosomes per Cell",
    fill = "Plate ID"
  )
```

```{r}
p.after <- df.ploidy.f %>%
  ggplot(aes( x = sample_id, y = ploidy, colour = as.factor(dilution_factor), shape = purpose)) +
  facet_wrap(~purpose, scales = "free_y", ncol = 1) +
  geom_jitter(width = 0.1) +
  coord_flip()

p.after
```

```{r}
(p.before  + labs(title = "Before") + theme(legend.position = 'none')) / (p.after + labs(title = "After") + theme(legend.position = 'none'))
```

```{r}

```

Another useful plot is to visualize every point per treatment.

```{r}

```

Last but not least, a plot showing whether there is a difference between the groups.

```{r}
df.ploidy.f %>% 
  group_by(sample_id, phosphate_mM) %>%
  summarise(
    ploidy = mean(ploidy)
  ) %>%
  ggplot(aes( x = as.factor(phosphate_mM), y = ploidy)) +
  stat_summary(fun.y = "mean", geom = "point", size=3) +
  geom_jitter(aes(color=sample_id)) +
  stat_summary(fun.data = "mean_sdl", fun.args = c(mult = 1), geom = "errorbar") +
  expand_limits(y=0)
```

Perform a t.test to see if there is a significant effect of phopshate concentration on the ploidy level.

```{r}
df.ploidy.f %>%
  group_by(sample_id, phosphate_mM) %>%
  summarise(
    ploidy = mean(ploidy)
  ) %>%
  t.test( ploidy ~ phosphate_mM, data = .)
```

```{r}
df.ploidy.f %>%
  group_by(sample_id, phosphate_mM) %>%
  summarise(
    ploidy = mean(ploidy)
  ) %>%
  wilcox.test( ploidy ~ phosphate_mM, data = .)
```

## Export

```{r}
ploidy.export.file.path <- df.ploidy.f %>% 
  select(sample_id, plate_id, dilution_factor, replicate, ploidy) %>%
  write_output_csv("ploidy", paste("chromosomes_per_cell", sep = "_"))
```

Data was exported to `r ploidy.export.file.path`

