---
title: "qPCR Analysis"
author: ""
date: ""
output: html_document
params:
  dataset:
    label: Dataset
    value: 200213_1
    input: select
    choices: [200207_1, 200207_2, 200212_1, 200213_1]
  processed:
    label: Processed
    value: F,
    input: select
    choices: [T, F]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(broom)
library(kableExtra)
library(patchwork)

theme_set(theme_bw(14))
```

```{r load_data}
# Load project settings
source(here::here("settings.R"))

# load output functions
source(file.path(functions.dir, "output.R"))

# parameters
plate.id <- params$dataset
# plate.id  <- "200212_1"
processed <- params$processed
# processed <- T

# load processed cell extract data
df.cell.extracts <- read_output_csv("cell_extracts", "cell_extracts")

# Load qpcr functions
source(file.path(data.raw.dir, "qpcr.R"))
df.plate.design <- load_plate_design_pcr(plate.id = plate.id)
df.protocol <- load_protocol_design_pcr(plate.id = plate.id)
df.qpcr.plate.raw <- load_export_pcr(plate.id = plate.id)
ncols.qpcr.plate <- ncol(df.qpcr.plate.raw)
```

```{r}
# Define background cycles 
bck.min <- 3
bck.max <- 7
# Define threshold
ct.threshold <- 0.2
# avogadro constant
avogadro <- 6.023E23 # particles / mol
# number of cycles for reaction stage
reaction.n.cycles <- 40
```

## Passive reporter normalization

Normalize the SYBR green signal with the ROX signal
```{r}
df.Rn <-  df.qpcr.plate.raw %>%
  rename(well = Well) %>%
  pivot_longer(cols = contains("Reading"), names_to = "cycle", values_to = "signal") %>% 
  pivot_wider(names_from = `Dye Name`, values_from = signal) %>% 
  mutate(
    cycle = as.numeric(str_remove(cycle, "Reading ")),
    # Calculate normalized reporter signal (Rn)
    Rn = SYBR / ROX
  )
```

## Background correction

Calculate the mean background Rn for each well

```{r df.qpcr.background}
## Get the average signal during the background cycles (defined on thi report)
df.qpcr.background <- df.Rn %>%
  group_by(well) %>%
  filter(cycle > bck.min & cycle < bck.max) %>%
  summarise(
    baseline = mean(Rn)
  )
```

Substract background from each reaction

```{r df.qpcr.corrected}
## Correct SYBR green signal with ROX and correct with baseline
df.qpcr.corrected <- df.Rn %>% 
  left_join(df.qpcr.background, by = "well") %>%
  group_by(well) %>%
  mutate(
    ARn = Rn - baseline
  ) %>%
  ungroup() %>%
  select(-c(ROX, SYBR))

```

```{r}
## Add design info to the corrected data
df.qpcr <- df.qpcr.corrected %>%
  left_join(df.plate.design, by = "well")  %>%
  # filter outliers if we are creating the processed document
  filter(!processed | !outlier)

# filter only reaction cycles
df.qpcr.reaction <- df.qpcr %>%
  filter(cycle < reaction.n.cycles) 
```

## Data inspection

### All cycles from assay (PCR reaction + melting curve)

```{r, fig.width=12, fig.height=6}
df.qpcr %>% 
  ggplot(aes(x = cycle, y = ARn, colour = sample_class, group = well)) +
  geom_line() +
  facet_grid(row~column) +
  labs(
    colour = ""
  ) +
  theme(legend.position = 'top')
```

### PCR reaction

```{r, fig.width=12, fig.height=7}
df.qpcr.reaction %>% 
  ggplot(aes(x = cycle, y = ARn, colour = as.factor(sample_class), group = well)) +
  geom_line() +
  facet_grid(row~column) +
  labs(
    x = "Cycle",
    y =  ~ paste(Delta, "Rn"),
    title = "Full plate", 
    colour = ""
  )+
  theme(legend.position = 'top')
```

#### Standards

```{r}
df.qpcr.reaction %>% 
  filter(sample_class == "standard") %>% 
  ggplot(aes(x = cycle, y = ARn, colour = as.factor(sample_id), group = well)) +
  scale_colour_viridis_d() +
  geom_line() +
  geom_hline(yintercept = ct.threshold) +
  facet_wrap(~replicate) +
  labs(
    x = "Cycle",
    y =  ~ paste(Delta, "Rn"),
    colour = "Sample",
    subtitle = "Black horizontal line shows the cutoff for Ct"
  ) 
```

```{r}
df.qpcr.reaction %>% 
  filter(sample_class == "standard") %>% 
  ggplot(aes(x = cycle, y = ARn, colour = as.factor(sample_id), group = well)) +
  scale_colour_viridis_d() +
  geom_line() +
  geom_hline(yintercept = ct.threshold) +
  facet_wrap(~sample_id) +
  labs(
    x = "Cycle",
    y =  ~ paste(Delta, "Rn"),
    colour = "Standard",
    subtitle = "Black horizontal line shows the cutoff for Ct"
  ) 
```

#### Samples

```{r}
df.qpcr.reaction %>% 
  filter(sample_class == "sample") %>% 
  ggplot(aes(x = cycle, y = ARn, colour = as.factor(dilution_factor), group = well)) +
  scale_colour_viridis_d() +
  geom_line() +
  geom_hline(yintercept = ct.threshold) +
  facet_wrap(~sample_id) +
  labs(
    x = "Cycle",
    y =  ~ paste(Delta, "Rn"),
    colour = "Dilution factor",
    subtitle = "Black horizontal line shows the cutoff for Ct"
  ) 
```

```{r}
df.qpcr.reaction %>% 
  filter(sample_class == "sample") %>% 
  ggplot(aes(x = cycle, y = ARn, colour = as.factor(replicate), group = well)) +
  scale_colour_viridis_d() +
  geom_line() +
  geom_hline(yintercept = ct.threshold) +
  facet_grid(dilution_factor~sample_id) +
  labs(
    x = "Cycle",
    y =  ~ paste(Delta, "Rn"),
    colour = "Sample",
    subtitle = "Black horizontal line shows the cutoff for Ct"
  ) 
```

#### NTC 

```{r}
df.qpcr.reaction %>% 
  filter(sample_class == "NTC") %>% 
  ggplot(aes(x = cycle, y = ARn, group = well)) +
  scale_colour_viridis_d() +
  geom_line() +
  geom_hline(yintercept = ct.threshold) +
  labs(
    x = "Cycle",
    y =  ~ paste(Delta, "Rn"),
    colour = "Dilution factor",
    subtitle = "Black horizontal line shows the cutoff for Ct"
  ) 
```

### Melting curve

```{r rt.pcr.melting.df}
# filter only melting curve
rt.pcr.melting.df <-  df.qpcr %>%
  filter(cycle > reaction.n.cycles) %>%
  group_by(well) %>%
  mutate(
    derivative = abs(c(NA, diff(Rn)))
  )

```

```{r}
rt.pcr.melting.df %>%
  ggplot(aes(x = cycle, y = ARn, group = well, colour = as.factor(sample_id))) +
  geom_line() +
  scale_colour_viridis_d() +
  theme(legend.position = "none") +
  expand_limits(x = 40) +
  facet_wrap(~sample_class) +
  labs(
    x = "Cycle",
    y =  ~ paste(Delta, "Rn"),
    colour = ""
  ) 
```

```{r}
rt.pcr.melting.df %>%
  filter(cycle > 60 ) %>% 
  ggplot(aes(x = cycle, y = derivative, group = well, colour = as.factor(sample_id))) +
  geom_line() +
  scale_colour_viridis_d() +
  expand_limits(x = 40) +
  facet_wrap(~sample_class) +
  theme(legend.position = "bottom") +
  labs(
    x = "Cycle",
    y =  ~ paste("-derivative", Delta, "Rn"),
    colour = ""
  )
```

### Ct counts

```{r}
df.qpcr.reaction %>% 
  filter(sample_class != "NTC") %>%
  ggplot(aes(x = cycle, y = ARn, group = well)) +
  scale_colour_viridis_d() +
  geom_line() +
  geom_hline(yintercept = ct.threshold) +
  facet_wrap(~sample_class) +
  scale_y_continuous(trans = "log10") +
  labs(
    x = "Cycle",
    y =  ~ paste("log10 ", Delta, "Rn"),
    colour = "Sample",
    subtitle = "Black horizontal line shows the cutoff for Ct"
  ) 
```

## Amplification efficiency

### Estimated slope, R^2^ and efficiency 

```{r ct.pcr.df}
# get the Ct value for each reaction
ct.pcr.df <- df.qpcr.reaction %>%
  filter(ARn <= ct.threshold) %>%
  group_by(well) %>%
  filter(cycle == max(cycle)) %>%
  rename(Ct = cycle) %>% 
  # attach info from standard fragment
  crossing(df.protocol %>% select(st_info) %>% unnest(st_info)) %>% 
  crossing(df.protocol %>% select(mm) %>% unnest(mm))
```


* Optimal slope: -3.32
* Recomendend R~2~ > 0.98
* Efficiency: 90-105 %

### Standard curve

```{r}
standard.curve.df <- ct.pcr.df %>%
  filter(sample_class == "standard") %>%
  mutate(
    well_conc = (initial_conc / dilution_factor) * template_vol / final_vol,
    # calculate # fragments per uL 
    fragments_uL = (well_conc / 1E9 /  amplicon_mw * avogadro)
  )

standard.model.df <- standard.curve.df %>%
  group_by(replicate) %>%
  mutate(fragments_uL = log10(fragments_uL)) %>%
  nest() %>%
  mutate(
    fit = map(data, ~lm(Ct ~ fragments_uL, data = .x))
  ) %>% 
  # add a standard curve including all the replicates
  bind_rows(
    standard.curve.df %>%
      ungroup() %>%
      mutate(
        replicate = "all",
        fragments_uL = log10(fragments_uL)
      ) %>%
      group_by(replicate) %>%
      nest() %>%
      mutate(
        fit = map(data, ~lm(Ct ~ fragments_uL, data = .x))
      )
  )

slopes.df <- standard.model.df %>%
  mutate(
    tidied = map(fit, tidy)
  ) %>%
  unnest(tidied) %>%
  filter(term == "fragments_uL") %>%
  select(replicate, slope = estimate)

r.squared.df <- standard.model.df %>%
  mutate(
    tidied = map(fit, glance)
  ) %>%
  unnest(tidied) %>%
  select(replicate, r.squared)

quality.df <- left_join(slopes.df, r.squared.df, by = "replicate") %>%
  mutate(
    E = 10^(-1/slope),
    efficiency = (E - 1) * 100
  )

```

```{r}
df.standard.model <- standard.model.df %>% 
  ungroup() %>%
  mutate(
    r.squared = map_dbl(fit, function(.d) { .d %>% glance %>% pull('r.squared')}),
    intercept = map_dbl(fit, function(.d) { .d %>% tidy %>% filter(term == "(Intercept)") %>% pull('estimate')}),
    slope = map_dbl(fit, function(.d) { .d %>% tidy %>% filter(term != "(Intercept)") %>% pull('estimate')}),
  ) %>%
  filter(r.squared > 0.98) %>%
  mutate( optimal_distance = abs(3.32 - abs(slope))  ) %>%
  arrange(optimal_distance) %>%
  slice(1)

standard.slope <- df.standard.model %>% pull(slope)
standard.int <- df.standard.model %>% pull(intercept)

standard.range.df <- standard.curve.df %>%
  ungroup() %>%
  summarise(
    max_Ct = max(Ct),
    min_Ct = min(Ct)
  )
```

Show the slope and intercept from the best standard curve

```{r}
df.standard.model %>% select(-fit, -data) %>%
  kable(format = "html") %>% 
   kable_styling(bootstrap_options = c("striped", "hover"))
```


```{r}
p.top <- standard.curve.df %>% 
  bind_rows(
    standard.curve.df %>%
      ungroup() %>% 
      mutate(
        replicate = "all",
        fragments_uL = fragments_uL
      )
  ) %>% 
  ggplot(aes(x = log10(fragments_uL), Ct)) +
  geom_point() +
  facet_wrap(~replicate, nrow = 1) +
  geom_smooth(method = "lm") +
  labs(
    x = "Log10 [DNA]",
    y = "Ct",
    title = "Standard curve"
  ) 

p.slope <- quality.df %>% 
  ggplot(aes(replicate, slope)) +
  geom_hline(yintercept = -3.32, colour = "red", size = 1) +
  geom_point(size = 2) +
  labs(
    x = "Sample",
    y = "Slope",
    subtitle = "Red line indicates optimal slope"
  )

p.r.sq <- quality.df %>% 
  ggplot(aes(replicate, r.squared)) +
  geom_hline(yintercept = c(0.98), colour = "red", size = 1, alpha = 0.5) +
  geom_point(size = 2) +
  labs(
    x = "Sample",
    y = expression(R^2),
    subtitle = "Red line indicates minimum R2"
  )

p.eff <- quality.df %>% 
  ggplot(aes(replicate, efficiency)) +
  geom_hline(yintercept = c(90, 105), colour = "red", size = 1, alpha = 0.5) +
  geom_point(size = 2) +
  labs(
    x = "Sample",
    y = "Efficiency",
    subtitle = "Red lines indicate optimal range"
  )
```

```{r, fig.width=10, fig.height=7}
p.top + {
  p.slope + p.eff + p.r.sq
} +
  plot_layout(ncol = 1)
```

```{r}
kable(quality.df, "html") %>% 
 kable_styling(bootstrap_options = c("striped", "hover"))
```

### Samples

```{r quality_checks_samples, eval = F}

samples.model.df <-  ct.pcr.df %>%
  filter(sample_class == "sample") %>%
  ungroup() %>%
  group_by(sample_id, replicate) %>%
  mutate(
    dilution_factor = as.numeric(dilution_factor)
  ) %>%
  mutate(
    fragments_uL = log10(1000/dilution_factor)
  ) %>%
  nest() %>%
  mutate(
    fit = map(data, ~lm(Ct ~ fragments_uL, data = .x))
  )

slopes.samples.df <- samples.model.df %>%
  mutate(
    tidied = map(fit, tidy)
  ) %>%
  unnest(tidied) %>%
  filter(term == "fragments_uL") %>%
  select(sample_id, replicate, slope = estimate)


r.squared.samples.df <- samples.model.df %>%
  mutate(
    tidied = map(fit, glance)
  ) %>%
  unnest(tidied) %>%
  select(sample_id, replicate, r.squared)

quality.samples.df <- left_join(slopes.samples.df, r.squared.samples.df, by = c("sample_id", "replicate")) %>%
  mutate(
    E = 10^(-1/slope),
    efficiency = (E - 1) * 100
  ) %>%
  arrange(sample_id, replicate)


```

Here I show the Ct values obtained for the samples. The vertical red lines inidicate the maximum and minimum Ct values obtained for the points in the standard curve. 

```{r}
ct.pcr.df %>% 
  filter(sample_class == "sample")  %>%  
  ggplot(aes(x = sample_id, y = Ct, colour = sample_dilution)) +
  geom_point(size = 2) +
  scale_colour_viridis_d() +
  geom_hline(yintercept =  standard.range.df$max_Ct, colour = "red") +
  geom_hline(yintercept =  standard.range.df$min_Ct, colour = "red") +
  facet_wrap(~replicate) +
  coord_flip() +
  labs(
    x = "Sample",
    y = "Ct",
    colour = "Sample's dilution"
  ) +
  theme(legend.position = "top")
```


```{r, fig.height=6, fig.width=10, eval = F}
theme_set(theme_bw(11))
p.st.s <- ct.pcr.df %>%
  filter(sample_class == "sample") %>%
  ungroup() %>%
  group_by(sample_id, replicate) %>%
  mutate(
    dilution_factor = as.numeric(dilution_factor)
  ) %>%
  mutate(
    fragments_uL = log10(1000/dilution_factor)
  ) %>% 
  ggplot(aes(x = fragments_uL, Ct, colour = replicate)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~sample_id, nrow = 1) +
  geom_smooth(method = "lm", alpha = 0.5) +
  labs(
    x = "Log10 [DNA]",
    y = "Ct",
    colour = "Replicate"
  ) +
  theme(legend.position = "top")


p.eff.s <- quality.samples.df %>% 
  ggplot(aes(sample_id, efficiency, colour = replicate)) +
  geom_hline(yintercept = c(90, 105), colour = "red", size = 1, alpha = 0.5) +
  geom_point(size = 2, alpha = 0.6) +
  labs(
    x = "Sample",
    y = "Efficiency",
    subtitle = "Red lines indicate optimal range"
  ) +
  theme(legend.position = "none")

p.rsq.s <- r.squared.samples.df %>% 
 ggplot(aes(sample_id, r.squared, colour = replicate)) +
  geom_hline(yintercept = c(0.98), colour = "red", size = 1, alpha = 0.5) +
  geom_point(size = 2, alpha = 0.6) +
  labs(
    x = "Sample",
    y = expression(R^2),
    subtitle = "Red line indicates minimum R2"
  ) +
  theme(legend.position = "none")
  
p.slope.s <- quality.samples.df %>% 
  ggplot(aes(sample_id, slope, colour = replicate)) +
  geom_hline(yintercept = -3.32, colour = "red", size = 1) +
  geom_point(size = 2, alpha = 0.6) +
  labs(
    x = "Sample",
    y = "Slope",
    subtitle = "Red line indicates optimal slope"
  )  +
  theme(legend.position = "none")
  
# generate plot
p.st.s + {
  p.slope.s + p.eff.s +  p.rsq.s
} +
  plot_layout(ncol = 1)

```


#### Distribution of number of DNA fragments per samples

```{r df.fragments}
df.fragments <- ct.pcr.df %>%
  filter(sample_class == "sample", Ct >= standard.range.df$min_Ct & Ct <= standard.range.df$max_Ct) %>%
  ungroup() %>%
  group_by(sample_id, replicate) %>%
  mutate(
    dilution_factor = as.numeric(dilution_factor)
  ) %>% 
  group_by(sample_id, replicate) %>%
  mutate(
    fragments_well_uL = 10^((Ct - standard.int) / standard.slope),
    fragments_dilution_uL = fragments_well_uL * final_vol / template_vol,
    fragments_sample_uL = dilution_factor * fragments_dilution_uL
  ) 
```

```{r}
df.fragments %>% 
  ggplot(aes(sample_id, fragments_sample_uL)) +
  geom_boxplot() +
  geom_jitter(aes(colour = as.factor(dilution_factor)), size = 2, alpha = 0.6, width=0.15, height = 0) +
  labs(
    x = "Sample",
    y = "[# DNA fragments]",
    colour = "Dilution factor"
  ) +
  coord_flip()
```

## Relationship of # DNA fragments vs dilution factor

```{r, fig.width=14, fig.height=6}
df.fragments %>% 
  ggplot(aes(dilution_factor, fragments_sample_uL, colour = sample_dilution)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  facet_grid(replicate~sample_id, scales = "free") +
  labs(
    x = "Dilution factor (log10)",
    y = "# DNA fragments"
  )
```

## Export

```{r}
if (processed) {
  qpcr.export.file.path <- df.fragments %>% 
    filter(sample_class == "sample") %>% 
    mutate(plate_id = plate.id) %>%
    select(plate_id, sample_id, replicate, dilution_factor, sample_dilution, fragments_sample_uL) %>%
    write_output_csv("qpcr", paste("qpcr_fragments", plate.id, sep = "_"))
  
  print(glue::glue("Data was exported to {qpcr.export.file.path}"))
} else {
  print("This document shows the unfiltered data so no data was exported.")
}
```
