---
title: "Metabolomics Analysis"
author: "Joeri Jongbloets"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning=FALSE
  
)

library("tidyverse")
library("broom")
# install.packages("ggfortify")
library("ggfortify")
# install.packages("colorspace")
library("colorspace")
```

## Exercise 1: Preparation

In the `load_data` chunk we need to load all the data we will need for the analysis. We have 2 data sources:

* `design`: maps sample_ids to the conditions of the experiment
* `metabolomics`: contains the measured intensities of each feature and mapping of metabolomics samples to sample_id.

The data is stored respectively in 3 data.frames: `df.design`, `df.metabolomics.raw` and `df.metabolomics.design`.



```{r load_data, include=FALSE}
# Load project settings
source(here::here("settings.R"))

# load output functions
source(file.path(functions.dir, "output.R"))
# Load pca functions
source(file.path(functions.dir, "pca.R"))

# Load experimental design
source(file.path(data.raw.dir, "design.R"))
df.design <- load_design()

# Load metabolomics data
source(file.path(data.raw.dir, "metabolomics.R"))
df.metabolomics.design <- load_metabolomics_design()
df.metabolomics.raw <- load_metabolomics()

df.metabolomics.buckets <- df.metabolomics.raw %>%
  distinct(phase, bucket, RT, mz_ratio, mass, charge, Name, Formula)

```


### Exercise 1.1

First thing to do when encountering a new dataset is to look at it's shape and properties.

```{r}
df.metabolomics.raw %>% glimpse()
```


#### Question 1.1

What is in data.frame and what do you think is the meaning of each column?
Is the data in a tidy format?



## Feature Statistics

As you might have notice, some of the features have a formula and/or a label associated to it but other don't. In addition, there are features that have the same formula and/or label. Create a figure showing the percentage of features with a formula and with a name for each phase and a table showing the total number of features and with names and formulas and how many of those are unique for each phase.

```{r}
df.metabolomics.summary <- df.metabolomics.buckets %>%
  group_by(phase) %>%
  summarise(
    # count total number of entries
    total_number = n(),
    # count number of entries with a formula. Hint: you can use `sum()` in combination with is.na()
    entries_with_formula = sum(!is.na(Formula)),
    # count number of unique formulas. Hint: `n_distinct` function
    unique_formula = n_distinct(Formula, na.rm = TRUE),
    # count number of entries with a name Hint: you can use `sum()` in combination with is.na()
    entries_with_name = sum(!is.na(Name)),
    # count number of unique names. Hint: `n_distinct` function
    unique_name = n_distinct(Name, na.rm = TRUE)
  )
```

```{r}
df.metabolomics.raw %>%
  group_by(phase, bucket) %>%
  summarise(
    n_samples = n_distinct(metabolomics_id)
  ) %>%
  ggplot(aes( n_samples )) +
  facet_wrap(~phase) +
  geom_histogram()
```


```{r}
df.metabolomics.summary %>%
  pivot_longer(
    cols = starts_with("entries_with_"),
    names_to = "type",
    names_prefix = "entries_with_",
    values_to = "counts"
  ) %>%
  mutate(
    # calculate the percentage of the total
    percentage = counts / total_number
  ) %>%
  ggplot(aes( x = type, y = percentage, fill = phase )) +
  # Hint: use position = "dodge" inside the geom_col to have side-by-side columns
  geom_col(position = "dodge") +
  # format labels on y-axis as percentages
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Variable",
    y = "% non-NA entries"
  )
```

```{r}
df.metabolomics.summary %>% 
  knitr::kable()
```

## Pre-Normalization

```{r}
df.metabolomics.raw %>%
  ggplot(aes( x = metabolomics_id, y = log10(value + 1), colour = phase )) +
  # facet_wrap(~phase) +
  geom_boxplot() +
  coord_flip() +
  labs(
    x = "Metabolomics ID",
    y = p.label.log("Value")
  )
```

```{r}
df.metabolomics.raw %>%
  left_join(df.metabolomics.design, by = c("phase", "metabolomics_id")) %>%
  ggplot(aes(log10(value + 1), colour = sample_id )) +
  facet_wrap(~phase) +
  geom_density() +
  labs(
    x = p.label.log("Value"),
    y = "Density",
    colour = "Metabolomics ID"
  )
```

## Normalize

In class we designed the following normalization method:

```{r}
df.metabolomics.norm <- df.metabolomics.raw %>%
  group_by(phase, bucket) %>%
  # remove metabolites with all zero values
  filter(sum(value) != 0) %>%
  left_join(df.metabolomics.design, by = c("phase", "metabolomics_id")) %>%
  left_join(df.design %>% select(sample_id, purpose), by = c("sample_id")) %>%
  group_by(phase, bucket, purpose) %>%
  mutate(
    expected_value = mean(value[value > 0]),
    expected_value = ifelse(is.na(expected_value), 0, expected_value),
    # remove/replace zero values with mean
    value_cor = ifelse(value > 0, value, expected_value)
  ) %>%
  group_by(phase, bucket) %>%
  mutate(
    # calculate pseudo-reference for size factors
    pseudo_reference = ifelse(any(value_cor == 0), NA, mean(log(value_cor)))
  ) %>%
  group_by(phase, metabolomics_id) %>%
  mutate(
    # calculate average distance to pseudo_reference
    size_factor = exp(median(log(value_cor) - pseudo_reference, na.rm=T)),
    # calculate normalized value
    value_norm = value_cor / size_factor,
  )  %>% select(-c(sample_id, purpose))
# NOTE: at this point the data.frame is grouped by phase AND metabolomics ID
```

Note: in certain situations (i.e. a metabolite with 0 values in one group) the above method will not replace these zeros.
Therefore you need to be aware of zero values in the rest of your analyasis.

The chunck below (note; eval = FALSE means it is not executed when knitted) demonstrates a different approach. 
This will replace zero values with the lowest observed value (per metabolite).

```{r, eval=FALSE}
df.metabolomics.norm <- df.metabolomics.raw %>%
  group_by(phase, bucket) %>%
  # remove metabolites with all zero values
  filter(sum(value) != 0) %>%
  mutate(
    # replace zero values by half of the smallest observed value for that metabolite
    value_norm = ifelse(value <= 0, min(value[value > 0], na.rm = T) / 2, value),
    # calculate pseudo-reference for size factors
    pseudo_reference = ifelse(any(value == 0), NA, mean(log(value)))
  ) %>%
  group_by(phase, metabolomics_id) %>%
  mutate(
    # calculate average distance to pseudo_reference
    size_factor = exp(median(log(value) - pseudo_reference, na.rm=T)),
    # calculate normalized value
    value_norm = value / size_factor,
  )
# NOTE: at this point the data.frame is grouped by phase AND metabolomics ID
```

### Check Normalization

```{r}
df.metabolomics.norm %>%
  ggplot(aes( x = metabolomics_id, y = log10(value_norm + 1), colour = phase )) +
  # facet_wrap(~phase) +
  geom_boxplot() +
  coord_flip() +
  labs(
    x = "Metabolomics ID",
    y = p.label.log("Value")
  )
```

```{r}
df.metabolomics.norm %>%
  left_join(df.metabolomics.design, by = c("phase", "metabolomics_id")) %>%
  ggplot(aes(log10(value_norm + 1), colour = sample_id )) +
  facet_wrap(~phase) +
  geom_density() +
  labs(
    x = p.label.log("Value"),
    y = "Density",
    colour = "Metabolomics ID"
  )
```

## PCA

```{r}
df.pca <- df.metabolomics.norm %>%
  group_by(phase) %>%
  # filter(phase == "polar") %>%
  select(phase, bucket, metabolomics_id, value_norm) %>% 
  # pivot_wider(
  #   names_from = all_of("bucket"), 
  #   values_from = all_of("value_norm")
  # )
  f_pca(counts_from = "value_norm", features_from = "bucket", samples_from = "metabolomics_id")
```

First look at the amount of variance explained by each component using the `f_pca_var_exp` function.

```{r}
df.pca.var.exp <- df.pca %>%
  f_pca_var_exp()

df.pca.var.exp
```

Or in plotted form using `f_pca_plot_var_exp`

```{r}
df.pca.var.exp %>%
  f_pca_plot_var_exp() +
  facet_grid(phase ~ key)
```

Which components are most relevant for explaining the variance in our dataset?

Let's inspect where our samples are placed on the first three components.

```{r}
df.pca %>%
  filter(phase == "polar") %>%
  mutate(
    pca_graph = map2(
      pca, data,
      f_pca_plot,
      x.pc = 1, y.pc = 2,
      design = df.metabolomics.design %>% 
        left_join(df.design, by = "sample_id"),
      design.key = "metabolomics_id",
      colour = "purpose", label = "sample_id"
    )
  ) %>% 
  pull(pca_graph) %>% first()
```

## Next steps:

From now on, you're on your own. Things to look at: 

- ~~Distribution of raw data~~
- Sample bias; ~~do you need to normalize?~~ Pick your normalization method
- PCA; interpret
- Heatmap
- Statistical tests

Note: beware there are two phases in this dataset, you should treat them as different datasets.
