---
title: "Multi-Cultivator Analysis"
author: "Your Name"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning=FALSE
  
)

library("tidyverse")

nest <- nest_legacy
unnest <- unnest_legacy
```

```{r load_data, include=FALSE}
# Load project settings
source(here::here("settings.R"))

# load growth calculation function
source(file.path(functions.dir, "growth.R"))
# load output helper functions
source(file.path(functions.dir, "output.R"))

# Load Multi-Cultivator
source(file.path(data.raw.dir, "multicultivator.R"))
df.multicultivator.raw <- load_multicultivator()

# Load Design
source(file.path(data.raw.dir, "design.R"))
df.design <- load_design() %>%
  filter(dataset == "course")

```

## Inspect Raw Data

Inspect the Multi-Cultivator data.frame.

```{r}
df.multicultivator.raw %>% glimpse()
```

Which columns are needed to calculate growth rates?

- multicultivator
- channel
- pump // decision
- time_h // time
- OD

### Optical Density in Multi-Cultivator

Show plot of the Optical Density over time.

```{r}
df.multicultivator.raw %>%
  ggplot(aes(x=time_h, y=OD, colour=pump)) + 
  facet_grid(multicultivator ~ channel) +
  geom_point() +
  labs(
    x = p.label.time.h,
    y = p.label.od.ln.720 
  )
```

Zoom-in on a few dilution cycles.

```{r}
df.multicultivator.raw %>%
  filter(time_h > 21.5, time_h < 29.5) %>%
  filter(multicultivator == 0, channel == 3) %>%
  ggplot(aes(x=time_h, y=OD, colour=pump)) + 
  facet_grid(multicultivator ~ channel) +
  geom_point() +
  labs(
    x = p.label.time.h,
    y = p.label.od.ln.720 
  )
```

Visualise the regressions for every dilution cycle.

```{r}
df.multicultivator.raw %>%
  filter(time_h > 21.5, time_h < 29.5) %>%
  filter(multicultivator == 0, channel == 3) %>%
  ggplot(aes(x=time_h, y=OD, colour=pump)) + 
  facet_grid(multicultivator ~ channel) +
  geom_point() +
  geom_smooth(aes(group=decision), se=F, method = "lm") +
  labs(
    x = p.label.time.h,
    y = p.label.od.ln.720 
  )
```

## Combine with Design

Our goal is to determine the growth rate of culture for every sample we took.
Therefore we need to associate our Multi-Cultivator data with the samples.

Can we do this using left join? Yes, we can BUT we have to careful with multicultivator, channel being a good predictor for sample_id.

```{r}
df.multicultivator.raw %>%
  left_join(
    df.design, by = c("multicultivator", "channel")
  ) %>%
  mutate(
    sample_time = ydm_hms(sample_time, tz="Europe/Amsterdam")
  ) %>%
  filter(time < sample_time)
```

```{r}
df.multicultivator <- df.design %>%
  mutate(
    sample_time = ydm_hms(sample_time, tz="Europe/Amsterdam")
  ) %>%
  select(sample_id, multicultivator, channel, sample_time) %>%
  mutate(
    data = pmap(
      list(m = multicultivator, c = channel, st = sample_time), function(m, c, st) {
        df.multicultivator.raw %>%
          filter(multicultivator == m, channel == c, time < st) %>%
          select(-c(multicultivator, channel))
      }
    )
  ) %>% 
  unnest(data)
```

Check if data from *after* sampling has been removed.

```{r}
df.multicultivator %>%
  ggplot(aes(x=time_h, y=OD, colour=pump)) + 
  facet_grid(multicultivator ~ channel) +
  geom_point() +
  labs(
    x = p.label.time.h,
    y = p.label.od.ln.720 
  )
```

```{r}
df.multicultivator %>%
  filter(multicultivator == 2, channel == 2) %>%
  filter(time_h > 10) %>%
  ggplot(aes(x=time_h, y=OD, colour=pump)) + 
  facet_grid(multicultivator ~ channel) +
  geom_point() +
  labs(
    x = p.label.time.h,
    y = p.label.od.ln.720 
  )
```

## Growth Rates

Calculate the growth rates on the smaller dataset using the `f_growth` function.
NOTE: to retain relevant group information, group by as many variables as possible.

```{r}
df.growth <- df.multicultivator %>%
  select(multicultivator, channel, sample_id, time_h, OD, decision) %>%
  group_by_at(vars(-c(time_h, OD, decision))) %>%
  f_growth()
```

Visualise all found growth rates.

```{r}
df.growth %>%
  ggplot(aes(x=time_h, y=slope, colour=r.squared)) +
  facet_wrap(~sample_id) +
  geom_point() +
  scale_color_viridis_c() +
  labs(
    x = p.label.time.h,
    y = p.label.mu,
    color = p.label.r.sq
  )
```

Can you guess 'correct' growth rate for our 'treatment' samples?
Which observations would you keep?

```{r}
df.growth %>%
  left_join(df.design, by="sample_id") %>%
  filter(purpose == 'control') %>%
  ggplot(aes(x=time_h, y=slope, colour=r.squared)) +
  facet_wrap(~sample_id) +
  geom_point() +
  scale_color_viridis_c() +
  labs(
    x = p.label.time.h,
    y = p.label.mu,
    color = p.label.r.sq
  )
```

Show a summary of the growth rates observed for each sample.

```{r}
df.growth %>%
  left_join(df.design, by="sample_id") %>%
  filter( purpose == 'control' ) %>%
  ggplot(aes(x=sample_id, y=slope, fill=sample_id)) +
  stat_summary(fun.y = "mean", geom="bar") +
  stat_summary(fun.data = "mean_sdl", geom="errorbar") +
  scale_fill_viridis_d() +
  coord_flip() +
  labs(
    x = "",
    y = p.label.mu,
    fill = ""
  )
```

We can try to improve our estimate by removing observation with low confidence (low R squared)

```{r}
df.growth %>%
  left_join(df.design, by="sample_id") %>%
  filter( purpose == 'control' ) %>%
  f_growth_filter() %>%
  ggplot(aes(x=sample_id, y=slope, fill=sample_id)) +
  stat_summary(fun.y = "mean", geom="bar") +
  stat_summary(fun.data = "mean_sdl", geom="errorbar") +
  scale_fill_viridis_d() +
  coord_flip() +
  labs(
    x = "",
    y = p.label.mu,
    fill = ""
  )
```

## Export

```{r export_data}
growth.rates.export.file.path <- df.growth %>%
  select(sample_id, decision, time_h, slope, r.squared, w_size) %>%
  write_output_csv("growth", "growth_rates")
```

Data was exported to `r growth.rates.export.file.path`
