---
title: "Growth Rate Analysis"
author: "Your Name"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning=FALSE
  
)

library("tidyverse")
```

```{r load_data, include=FALSE}
# Load project settings
source(here::here("settings.R"))

# load output helper functions
source(file.path(functions.dir, "growth.R"))
# load output helper functions
source(file.path(functions.dir, "output.R"))

# load design
source(file.path(data.raw.dir, "design.R"))
df.design <- load_design()

# load all processed growth rates
df.growth.rates <- read_output_csv("growth", "growth_rates(_pc_[0-9])?")
```

List all sample IDs that are contained in the dataset

```{r}
df.growth.rates %>%
  pull(sample_id) %>%
  unique()
```

Show all the collected growth rates

```{r}
df.growth.rates %>%
  left_join(df.design, by = "sample_id") %>%
  f_growth_filter() %>%
  ggplot(aes(x=sample_id, y=slope, fill=sample_id)) +
  stat_summary(fun.y = "mean", geom = "bar") +
  stat_summary(fun.data = "mean_sdl", geom = "errorbar") +
  scale_fill_viridis_d() +
  coord_flip()
```

You harvested while the culture was crashing (negative growth). We actually harvested from "steady"-states. So we *can* use the growth rates from those samples. Create a plot with only the growth rates from sample in steady state.
NOTE: you can use assume all pre-course samples are from steady state.

```{r}
df.growth.rates %>%
  left_join(df.design, by = "sample_id") %>%
  filter( purpose == 'control' | str_detect(sample_id, "_PC_")) %>%
  f_growth_filter() %>%
  ggplot(aes(x=sample_id, y=slope, fill=sample_id)) +
  stat_summary(fun.y = "mean", geom = "bar" ) +
  stat_summary(fun.data = "mean_sdl", geom="errorbar") +
  scale_fill_viridis_d() +
  coord_flip() +
  labs(
    x = "",
    y = p.label.mu,
    fill = ""
  )
```

```{r}
df.growth.rates %>%
  left_join(df.design, by = "sample_id") %>%
  filter( purpose == 'control' | str_detect(sample_id, "_PC_")) %>%
  f_growth_filter() %>%
  ggplot(aes(x=sample_id, y=slope, fill=purpose)) +
  stat_summary(fun.y = "mean", geom = "bar" ) +
  stat_summary(fun.data = "mean_sdl", geom="errorbar") +
  scale_fill_viridis_d() +
  coord_flip() +
  labs(
    x = "",
    y = p.label.mu,
    fill = ""
  )
```
