---
title: "CASY-counter"
author: "Hugo Pineda Hernandez"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning=FALSE, message = FALSE
  
)

library(tidyverse)

unnest <- unnest_legacy
nest <- nest_legacy

```

```{r load_data, include=FALSE}
# Load project settings
source(here::here("settings.R"))

# load output functions
source(file.path(functions.dir, "output.R"))

# Load Experimental Design
source(file.path(data.raw.dir, "design.R"))
df.design <- load_design()

# Load CASY Count Data
source(file.path(data.raw.dir, "casy.R"))
df.casy.design <- load_casy_design()
df.casy <- load_casy()
```

### Explore raw data

```{r}
df.casy %>% 
  ggplot(aes(x = size_um, y = counts, colour = replicate)) +
  geom_point(alpha = 0.5) +
  facet_grid(sample_type~sample_id) +
  labs(
    x = p.label.size,
    y = p.label.counts
  ) 
```


### Calculate # cells/mL

1. Sum all the counts reported by the CASY-counter for each sample and replicate.
2. Correct by the corresponding dilution factor.

```{r}
df.casy.counts <- df.casy %>%
  # group by all variables BUT counts and size_um
  group_by_at(vars(-c(counts, size_um))) %>% 
  summarise(
    total_counts = sum(counts)
  ) %>% 
  mutate(
    # first calculate concentration in casy counter (counts / 200 uL)
    casy_mL = total_counts / (measurement_volume / 1E3),
    # calculate dilution made in casyton
    casy_dilution = (sample_volume + casyton_volume) / sample_volume,
    # calculate final concentrations using all dilution steps
    cells_mL = casy_mL * casy_dilution * sample_dilution
  )
```

#### Technical replicates

Inspect variation between technical replicates for the counts.

```{r}
df.casy.counts %>% 
  ggplot(aes(x = sample_id, y = cells_mL, colour = replicate)) +
  geom_point() +
  facet_wrap(~sample_type, scales = "free") +
  coord_flip()
```

### Calculate average and peak diameter

```{r}
df.casy.diameter <- df.casy %>%
  group_by(sample_id, sample_type, replicate) %>%
  arrange(size_um) %>%
  summarise(
    mean_diameter = sum(size_um * counts) / sum(counts),
    median_diameter = size_um[which.min(abs(cumsum(counts) - sum(counts) / 2))],
    peak_diameter = size_um[which.max(counts)]
  )
```


### Join count with diameter data

```{r}
df.casy.full <- left_join(
  df.casy.counts,
  df.casy.diameter,
  by = c("sample_id", "sample_type", "replicate")
)
```

### Validation of peak diameter estimation 

```{r}
df.casy %>%
  left_join(df.casy.diameter, by=c("sample_id", "sample_type", "replicate")) %>%
  # to gain resolution in our plot we can filter to a specific subset
  # filter(sample_type == "MC", size_um < 5, sample_id == "200206_2_5") %>%
  pivot_longer(
    cols = c(peak_diameter, mean_diameter, median_diameter),
    names_to = "size_type",
    values_to = "diameter"
  ) %>%
  ggplot(aes(x=size_um, y=counts)) +
  geom_point(alpha = 0.1) +
  geom_vline(aes(xintercept = diameter, color = size_type)) +
  facet_grid(sample_id ~ replicate) +
  labs(
    x = p.label.size,
    colour = ""
  )
```

### Summarise data

```{r}
df.export <- df.casy.full %>%
  group_by(sample_id, sample_type) %>%
  summarise(
    cells_mL = mean(cells_mL),
    peak_diameter = mean(peak_diameter),
    mean_diameter = mean(mean_diameter),
    median_diameter = mean(median_diameter)
  )
```

### Export data

```{r export_data}
df.export %>%
  write_output_csv("casy", "casy_counts_diameter")
```

