---
title: "CASY-counter"
author: "Hugo Pineda Hernandez"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning=FALSE, message = FALSE
  
)

library(tidyverse)

unnest <- unnest_legacy
nest <- nest_legacy

```

```{r load_data, include=FALSE}
# Load project settings
source(here::here("settings.R"))

# load output functions
source(file.path(functions.dir, "output.R"))

# Load Experimental Design
source(file.path(data.raw.dir, "design.R"))
df.design <- load_design()

# Load CASY Count Data
source(file.path(data.raw.dir, "casy.R"))
df.casy.design <- load_casy_design()
df.casy <- load_casy()
```

### Explore raw data

```{r}
df.casy %>% 
  ggplot(aes(x = size_um, y = counts, colour = replicate)) +
  geom_point(alpha = 0.5) +
  facet_grid(sample_type~sample_id) +
  labs(
    x = p.label.size,
    y = p.label.counts
  ) 
```


### Calculate # cells/mL

1. Sum all the counts reported by the CASY-counter for each sample and replicate.
2. Correct by the corresponding dilution factor.

```{r}
df.casy.counts <- df.casy %>%
  # group by all variables BUT counts and size_um
  group_by_at(vars(-c(counts, size_um))) %>% 
  summarise(
    total_counts = sum(counts)
  ) %>% 
  mutate(
    # first calculate concentration in casy counter (counts / 200 uL)
    casy_mL = total_counts / measurement_volume,
    # calculate dilution made in casyton
    casy_dilution = (sample_volume + casyton_volume) / sample_volume,
    # calculate final concentrations using all dilution steps
    cells_mL = casy_mL * casy_dilution * sample_dilution
  )
```

#### Technical replicates

Inspect variation between technical replicates for the counts.

```{r}
df.casy.counts %>% 
  ggplot(aes(x = sample_id, y = cells_mL, colour = replicate)) +
  geom_point() +
  facet_wrap(~sample_type, scales = "free") +
  coord_flip()
```

### Calculate average and peak diameter

```{r}

```


### Join count with diameter data

```{r}

```

### Validation of peak diameter estimation 

```{r}

```


### Summarise data and join design

```{r}

```


### Export data


```{r export_Data}
df.export %>%
  write_output_csv("casy", "casy_counts_diameter")
```

